---
title: "Functions"
output: 
  learnr::tutorial:
    progressive: false
    css: "css/tutorial.css"
    theme: "sandstone"
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, error = TRUE, exercise.timelimit = 3)
```

## Introduction

---

### Learning objectives

- Function calls and syntax
- Writing functions
- Best practices

---

The ability to write and understand functions will elevate your skill set as a
data scientist, statistician, and R programmer. By writing and using functions
you can tackle one big problem by breaking it up into smaller key steps, where
a set of functions you created is used to complete the 
necessary tasks. Incorporating functions into your workflow will

- make your work more reproducible,
- make your code more readable,
- improve your coding style, and
- decrease the number of bugs in your code.

### Packages

The following packages have been loaded in this tutorial:

```{r}
library(tidyverse)
```

<br/>

Video to go here...

## Function calls

You'll be using functions a lot, from those available in R packages to ones you
write. Most functions perform some action(s) to some object(s). 
When using functions

- you can leave out the name of the data argument (generally this is first);
- write out the names of other arguments;
- do not write out the default arguments and their values unless you change the
  values.

```{r eval=FALSE}
# Good
mean(1:20)
ggplot(mtcars, mapping = aes(x = hp, y = mpg)) +
  geom_point()

# Bad
mean(1:20, trim = 0, na.rm = FALSE)
ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point()
```

<div class = "warning">
**Warning!** In fitting a linear model, function `lm()` has the data argument
second. Always check the help (`?lm`) when working with a new function to 
understand its arguments.
</div>

```{r help_lm, exercise=TRUE}
?lm
```

## Writing functions

Functions in programming and data science are not unlike the mathematical
functions you learned about in pre-calculus. Functions depend on a set of
inputs. You'll pass values into those inputs and based on the function's
purpose it will generally return a result.

### Function syntax

A function is comprised of arguments (formals), a body, and an environment. 
The first two will be our focus as we use and develop these objects.

Below is the syntax for writing your own function in R. 

```{r eval=FALSE}
fcn_name <- function(arg_1, arg_2) {

  # function's body
  # code that does the work goes here

}
```

- `fcn_name` is a name you choose for your function.
- `function` is a keyword in R that you will use when you create your function.
- `arg_1`, `arg_2` are the function's arguments. You choose how many, their
  names, and their purpose. Keep the names short and descriptive.
- Between the `{` `}` will be the R code for your function's body. Generally,
  this code will manipulate data based on the values input to the arguments and
  then return a result.

### Exercise 1 {.tabset .tabset-fade .tabset-pills}

Take a look at the simple function below. Can you identify the function's 
name, its arguments, and its body?

#### Function

```{r ex_1, exercise=TRUE}
preview_data <- function(data) {
  data %>% 
    sample_n(size = 5)
}
```

#### Solution

| Function component           | Value               |
|:-----------------------------|:--------------------|
| Name                         | `preview_data`      |
| Argument                     | `data`              |
| Contents                     | Code on lines 2 - 3 |

### Function call

Run the code below to see this function in action. We'll always need to load
the function into our environment before using it. This is as simple as putting
it in an R chunk in your R Markdown document and hitting the green play button.

```{r preview_data, exercise=TRUE, error=TRUE}
preview_data <- function(data) {
  data %>% 
    sample_n(size = 5)
}

preview_data(data = mtcars)
```

<div class = "info">
**Note!** Here we have `data = mtcars` to be explicit. Typically you would
make your function call by `preview_data(mtcars)` as the data argument is
typically first.
</div>

R automatically returns the result of the last evaluated expression inside a
function. Alternatively, you can use `return()` explicitly. For example, another
way to write `preview_data()` is

```{r}
preview_data <- function(data) {
  x <- data %>% 
    sample_n(size = 5)
    
  return(x)
}
```

### Exercise 2

Consider a function that takes in a vector of numeric data and returns a 
vector of length three that contains the mean, median, and standard deviation.
Fill in the missing parts (`---`) below to complete the function below. 
Once you've finished the function, run the code.


```{r get_stats, exercise=TRUE}
get_stats <- ---(x) {
  ---
}

data <- c(3, 0, 1, 5, -4, 10)
get_stats(x = data)
```

```{r get_stats-solution}
get_stats <- function(x) {
  c(mean(x), median(x), sd(x))
}

data <- c(3, 0, 1, 5, -4, 10)
get_stats(x = data)
```

### Exercise 2

Create a function called `pluck_col()` that will grab a column from a data 
frame. Test your new function on the `longley` data set available in R. As a 
reminder,`tidyverse` was loaded when this tutorial was launched.

```{r}
longley
```

```{r pluck, exercise=TRUE}
# build pluck_col here


pluck_col(longley, col = "Population")
```

```{r pluck-solution}
pluck_col <- function(data, col) {
  data %>% 
    select(-col)
}
```

## Best practices

Now that you know the syntax of a function in R and the basics, it's time to 
explore some best practices with regards to writing functions. When should you 
write a function? What should you name your function? Should you use `return()`?

### When should you write a function?

If you copy the same code more than twice, write a function. For example,
suppose you have vectors of numeric data you want to standardize. Rather than
writing this code

```{r standardize_code, exercise=TRUE}
x <- 1:10
y <- c(4, 0, -4)
z <- c(0.01, 0.1, 1, 10, 100, 1000)

(x - mean(x)) / sd(x)
(y - mean(y)) / sd(y)
(z - mean(z)) / sd(z)
```

write a function with a single data argument.

```{r standardize_function, exercise=TRUE}
standardize <- function(data) {
  (data - mean(data)) / sd(data)
}

x <- 1:10
y <- c(4, 0, -4)
z <- c(0.01, 0.1, 1, 10, 100, 1000)


standardize(x)
standardize(y)
standardize(z)
```

Writing functions will help make your code more readable, and you'll be less
prone to copy-paste mistakes. Also, if you ever need to revise your 
standardization procedure, you only need to do so in the function's body. This
is much more efficient.

### Naming

Functions generally perform some action to a set of data conditional on a
set of values passed to the function's arguments. When naming your function,
choose a short and memorable verb. Think about the `dplyr` functions you know:
`select()`, `filter()`, `group_by()`, `summarize()`. All are short, memorable,
and give some insight as to what the function does. Strive to accomplish
this when you name your functions.

Consider a set of functions you would create to scrape and save available jobs 
from an online job board.

```{r eval=FALSE}
# Good
get_jobs()
clean_jobs()
write_jobs()


# Bad
jobs_from_online_job_board()
jobs_cleaned()
saving_my_jobs()
```

### Using `return()`

A function written in R will automatically return the last expression evaluated.
It is recommended that you only use `return()` for early returns. This will
occur when you start writing more complex functions with control flow. For
now, rely on R to return the result.

```{r eval=FALSE}
# Good
is_letter <- function(l) {
  l %in% letters
}

# Bad
is_letter <- function(l) {
  return(l %in% letters)
}
```

## Check your understanding

Below are some questions to check your understanding on functions.

```{r quiz, echo=FALSE}
quiz(
  question("A user-created function can only have 1 or 2 arguments.",
    answer("True", 
           message = "Functions can have any number of arguments, 
                      including 0."),
    answer("False", 
           correct = TRUE, 
           message = "Functions can have any number of arguments, 
                      including 0."),
    random_answer_order = TRUE
  ),
  
  question("The R keyword `function` is used to create functions.",
    answer("True", 
           correct = TRUE,
           message = "This is one of the key components in the function
                      creation syntax."),
    answer("False", 
           message = "This is one of the key components in the function
                      creation syntax."),
    random_answer_order = TRUE
  ),
  
  question("In general, function names should be",
    answer("nouns.", 
           message = "Typically, functions perform some action."),
    answer("adjectives.",
           message = "Typically, functions perform some action."),
    answer("long and descriptive.",
           message = "Function names should be short and memorable."),
    answer("verbs.",
           correct = TRUE,
           message = "Use an action word if you can."),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  
  question("Functions that exist in your main environment or in loaded 
            packages can be used inside functions you create.",
    answer("True",
           correct = TRUE,
           message = "R has the convention of lexical scoping."),
    answer("False", 
           message = "In the example there was a `dplyr` function inside
                      `preview_data()`."),
    random_answer_order = TRUE
  ),
  
  question("Consider a function called `compute_area()`, where the first line
            in this user-defined function is 
            `compute_area <- function(radius = 1) {`. Argument `radius` has a 
            default value of 1.",
    answer("True",
           correct = TRUE,
           message = "Setting arguments equal to a value when the function
                      is created is the way to set default values."),
    answer("False",
           message = "This is how default values are set for function
                      arguments."),
    random_answer_order = TRUE
  ),
  
  question("The `{` and `}` wrap the function's body. Which statement about
            the braces is correct?",
    answer("The braces are not necessary. A single tab is enough to let R know
            the subsequent code corresponds to the function's body.",
           message = "Not having the braces will cause problems."),
    answer("Both braces go on the same line.",
           message = "This style is not recommended. You may see it for very
                      small functions."),
    answer("The opening brace is placed after the function is defined and ends
            the line. The closing brace is placed on a line by itself at the
            conclusion of the code for the function's body.",
           correct = TRUE,
           message = "This is the best style."),
    answer("The opening brace is placed on a new line after the function is
            defined. The closing brace is placed on a line by itself at the
            conclusion of the code for the function's body.",
           message = "The opening brace should end the first line in which
                      the function is defined."),
    answer("The braces can only wrap up to 80 lines of code in a 
            single function.",
           message = "The function's body has no line limit."),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
  
)
```

## References

1. Grolemund, G., & Wickham, H. (2020). 19 Functions | R for Data Science. 
   Retrieved from https://r4ds.had.co.nz/functions.html
   
2. Wickham, H. (2020). The tidyverse style guide. Retrieved from
   https://style.tidyverse.org/