---
title: "k nearest neighbor"
author: "Maria Tackett"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```

By the end of this tutorial, you be able to:

- Use `dplyr` to create training and test data sets
- Use R to conduct knn
- Compare actual and predicted classifications

The following packages are used in this tutorial:

```{r load-packages}
library(tidyverse)
library(class)
library(palmerpenguins)
```

## The Data

For this tutorial, we are using the `penguins` dataset from the [`palmerpenguins` R package](https://allisonhorst.github.io/palmerpenguins/index.html) maintained by Dr. Allison Horst. This data contains measurements and other characteristics for 344 penguins observed near Palmer Station Antartica. The data were originally collected by Dr. Kristen Gorman. Click [here](https://allisonhorst.github.io/palmerpenguins/reference/penguins.html) to view the codebook for this data set.

```{r}
glimpse(penguins)
```

There are some observations with missing values for the measurements, so we will remove those for the purposes of this analysis: 

```{r}
penguins <- penguins %>%
  drop_na()
```

## Exploratory data analysis

We will use k-nearest neighbor to classify the penguins into three groups based on their measurements. Before performing the classification, let's do a brief exploratory data analysis to take a look at the data. 

```{r}
ggplot(data = penguins, aes(x = species)) +
  geom_bar(fill = "cyan4") +
  labs(title = "Distribution of Penguin Species")
```

## k-nearest neighbor

We will use the `knn` function from the `class` package. Let's start by splitting the data into training and test sets using `dplyr` functions.

```{r}
set.seed(072020) #set a seed since we will use random sampling

# shuffle the observations
penguins_shuffle <- penguins %>%
  sample_n(nrow(penguins), replace = FALSE)

n <- nrow(penguins_shuffle)

train <- penguins_shuffle %>%
  slice(1:(n-10))

test <- penguins_shuffle %>%
  slice((n-9):nrow(penguins_shuffle))
```

```{r}
train_species <- train %>%
  select(species) %>%
  pull()

test_species <- test %>%
  select(species) %>%
  pull()
```

```{r}
train <- train %>%
  select(body_mass_g, ends_with("_mm"))

test <- test %>%
  select(body_mass_g, ends_with("_mm"))
```

Let's perform k-nearest neighbor
```{r}
library(class)

knn_species <- knn(train, test, train_species, k = 3, prob = FALSE)
```

Let's compare the actual values of species with the ones produced by knn for the penguins in the test set. 

```{r}
results <- tibble(actual = test_species, 
                  predicted = knn_species)

results %>%
  count(actual, predicted)
```



