---
title: "Spatial Data in R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction to Spatial Data
The demographer [Waldo Tobler](https://en.wikipedia.org/wiki/Waldo_R._Tobler) once wrote "Everything is related to everything else, but near things are more related than distant things." This idea became known as Tobler's first law of geography. What this means is that observations that are close to one another in space are likely to display similarities, while observations that are further away may be more different.

In this tutorial, we will discuss how to check for spatial dependencies and how to run spatial regression models. You will also learn how to make maps in R.

## What is a nearby observation?
In spatial modeling, there are different ways to define nearby observations. Sometimes you might look at whether an observation borders another. Spatial modelers sometimes use chess moves to define contiguity. Imagine your unit of geography as a giant chess board. A [rook](https://geodacenter.github.io/workbook/4a_contig_weights/lab4a.html#rook-contiguity) contiguous observation (to use one example) would be one that you could move to from that observation in a game of chess with a rook.

Other times you might select the k-nearest observations (regardless of whether they are next door). Still other times, you might be interested in looking at exactly how far away each observation is from that observation. 

## Checking for Spatial Dependency
The [spData](https://cran.r-project.org/web/packages/spData/spData.pdf) package has a number of spatial data sets built in that you can work with for practice. First, we are going to look at data on voter turnout in the 1980 election.

Based on Tobler's First Law, we would expect that counties with high voter turnout might be clustered closely with other counties with high turnout (and the same for low turnout counties).

But how do we know if two observations in space display a similarity on a variable that we are interested in examining? Two common indices of spatial autocorrelation are Moran's I and Geary's C. Moran's I is a global measure of spatial autocorrelation-- that is, it looks at the data set as a whole-- while Geary's C is more attuned to local autocorrelation. 

There are several packages in R that you can use to calculate these indices of spatial autocorrelation. here, we are going to focus on the spdep package. (Other packages you can use include the [ape](https://cran.r-project.org/web/packages/ape/ape.pdf) package.

First, load spDataLarge using the below code.
```
install.packages('spDataLarge',
repos='https://nowosad.github.io/drat/', type='source')
```

Then, you are ready to calculate Moran's I and Geary's C. As you can see below, there is a significant amount of spatial autocorrelation in our dependent variable here for both indices.

```{r spatial, echo=TRUE}
library(spdep)
coordinates <- as.matrix((cbind(elect80$long, elect80$lat)))
coord_data.set<-coordinates(coordinates)
nearestneighbors <-knn2nb(knearneigh(coord_data.set, k=10))
weight <- nb2listw(nearestneighbors, style="W")
moran.test(elect80$pc_turnout, weight)
geary.test(elect80$pc_turnout, weight)

```

## Running a Spatial Regression Model
Now that we've determined that there are some spatial dependencies, we need to account for them when running a regression model. As with finding indices of spatial autocorrelation, there are several ways to do this in R. Here, we will use the spatialreg package.

Let's say that you think that voter turnout is a function of education (percent of people in a county with college degrees), homeownership, and income. You would run a model with each of these as independent variables and voter turnout as the dependent variable. 

There are various types of spatial regression [models](https://msarrias.weebly.com/uploads/3/7/7/8/37783629/lecture2.pdf), including spatial lag, spatial error, and Spatial Durbin models. 

Here, we are going to run a spatial lag model. These models include a lagged term ρ (rho) that accounts for the effect that observations have on nearby observations. In this context, this would account for the effect that high turnout has on nearby counties.

To run the spatial lag model, we use the lagsarlm command in spatialreg. (If we wanted to run a spatial error model, we would use errorsarlm.) Here, we list our model, the data, and then our weight matrix. Here, we are using the nearest neighbor matrix from earlier. In other cases, you might think that, for example, an inverse distance matrix might better reflect the process of spatial spillover effects, so you might construct that matrix here instead.

Using the summary command, we can see that each of these factors is statistically significant. Additionally, ρ is also relatively large and significant, reflecting the spatial autocorrelation present in the dependent variable.

```{r spatial model, echo=TRUE}
library(spatialreg)
spatialmodel1<- lagsarlm(pc_turnout~pc_college +pc_income+pc_homeownership, data=elect80, listw=weight)
summary(spatialmodel1)
```

## Using R to Visualize Spatial Data Into R
So far, we've used R to check for spatial autocorrelation and to run a spatial regression model. Another cool thing that you can do with spatial modelling is make maps in R. As with everything so far, there are [multiple ways](https://rspatial.org/raster/spatial/9-maps.html) to make plots, but here we are going to focus on using spplot and raster. The first thing that you have to do is load the [shapefile](https://doc.arcgis.com/en/arcgis-online/reference/shapefiles.htm), which is a file that stores geographic information such as location. 

For our purposes here, the spData packages has shapefiles ready to load in. The one we are going to use here comes from the Columbus data set, which has variables such as the crime rate and income for each of the 49 neighborhoods in Columbus, Ohio. Here, we load in the shapefile (.shp) and then plot the base map.
```{r plot, echo=TRUE, fig.width=5}
library(raster)
p <- shapefile(system.file("shapes/columbus.shp", package="spData"))
plot(p)
```

Once you have the base map, there are plenty of things that you can do to customize your map. Here I add a label for neighbor number using text() and you can color in each neighborhood a different color using rainbow(n). 

```{r plot2, echo=TRUE, fig.width=5}
n <- length(p)
plot(p, col=rainbow(n))
text(p, columbus$NEIG)
```


If you would like to make a somewhat more complicated plot then you might want to use spplot. Here, I use spplot to shade in the neighborhoods by income level and added an overall title.

```{r spplot}
spplot(p, "INC", main="Columbus Neighborhood Income in 1980")
```

## Some Notes on Loading Your Own Spatial Datasets
Here, we used spatial data that was already in R. But what if you want to load in your own data?

There are a few things to remember here. First, you are going to need to create a spatial weights matrix. If you want to look at the k-nearest neighbors (based on distance) or create an inverse weights matrix then having coordinates (e.g., longitude and latitude) may suffice. If you instead want to use some form of contiguity, then you can use a [.GAL file]( .https://cran.r-project.org/web/packages/spdep/vignettes/nb.pdf that indicate). 

Finally, if you want to do mapping in R, then you will need to have a shape file (.shp).

### Quiz

Here are two quick quiz questions to wrap up this tutorial:

```{r quiz}
quiz(
  question("Which of these is an index of autocorrelation? Select all that apply.",
    answer("Durbin's Z"),
    answer("Geary's C", correct=TRUE),
    answer("Weinberg's Q"),
    answer("Moran's I", correct=TRUE)
  ),
  question("Which kind of file do you need to make a map in R with plot/spplot?",
    answer(".GAL"),
    answer(".SHP", correct=TRUE),
    answer(".dta"),
    answer(".nb2")
  )
)
```

