---
title: "Automation"
output:
  learnr::tutorial:
    progressive: no
    theme: cerulean
    allow_skip: yes
runtime: shiny_prerendered
resource_files:
- automation.Rmd
- automation.Rmd
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, error = TRUE, exercise.timelimit = 5)
```

## Introduction

---

### Learning objectives

- When to use iteration
- Understand the `for` loop structure
- When to not use a loop

---

Writing and using functions is one way to reduce duplication in your code.
However, functions alone will not be able to eliminate all instances of 
duplication. You may be able to write a function to scale each numeric variable
in a data frame to a [0, 1] scale. But, if you need to apply this function
to 30 variables in that data frame there will be duplication. This is where
iteration (automation) plays a role in reducing duplication. R offers loops
and functionals as tools for reducing duplication through iteration. This
tutorial will introduce you to automation through `for` loops.

<div class="info">
**Note!** A functional is a function that takes a function as its argument and
returns a vector as output.
</div>

### Packages

The following packages have been loaded in this tutorial:

```{r}
library(tidyverse)
library(nycflights13)
```

## Why automation?

Consider the small tibble below.

```{r include=FALSE}
x <- tibble(
  col_a = c(3, -1, 0, 10),
  col_b = c(2, -2, 2, -22),
  col_c = c(8, sqrt(131), log(4), 33),
  col_d = 1:4
)
```


```{r example_tibble}
x <- tibble(
  col_a = c(3, -1, 0, 10),
  col_b = c(2, -2, 2, -22),
  col_c = c(8, sqrt(131), log(4), 33),
  col_d = 1:4
)

x
```

Suppose we were interested in computing the mean of each column. One technique
may be as follows.
 
```{r pull_1}
x %>% 
  pull(1) %>% 
  mean()
```
 
```{r pull_2}
x %>% 
  pull(2) %>% 
  mean()
```

```{r pull_3}
x %>% 
  pull(3) %>% 
  mean()
```

```{r pull_4}
x %>% 
  pull(4) %>% 
  mean()
```

Unfortunately, this results in 12 lines of code with a lot of duplication. 
A slightly better technique would be to first create a function and then use
that function.

```{r include=FALSE}
get_column_mean <- function(data, index) {
  data %>% 
    pull(index) %>% 
    mean()
}
```

```{r get_column_mean_fcn, exercise=TRUE}
get_column_mean <- function(data, index) {
  data %>% 
    pull(index) %>% 
    mean()
}
```

To get the column means of `x`, we can now use the following.

```{r run_function, exercise=TRUE}
get_column_mean(data = x, index = 1)
get_column_mean(data = x, index = 2)
get_column_mean(data = x, index = 3)
get_column_mean(data = x, index = 4)
```

We've reduced the number of lines of code from 12 to nine. However, there still
exists four lines of nearly the same code. All that varies is the column index
we use to compute the mean. This works okay here since we only have four 
columns in our data frame. What if
we had 100, or 1,000, or even 10,000 columns? Data this large is not uncommon, 
and we certainly do not want to have 10,000 function calls of 
`get_column_mean()` where the only variation is in the index.

How can we reduce the code duplication even more?

## Iteration with `for` loops

Looking back at our previous code, we see that the only variation is with
regards to the column index in which we want to compute the mean.

```{r eval=FALSE}
get_column_mean(data = x, index = 1)
get_column_mean(data = x, index = 2)
get_column_mean(data = x, index = 3)
get_column_mean(data = x, index = 4)
```

A `for` loop
is one type of loop available in R that can easily automate our process and
eliminate this duplication.

<div class="info">
**Note!** R also has `while` and `repeat` loops. The `for` loop is the most
common and is available in almost every programming language. More information
on `while` loops can be found
[here](https://r4ds.had.co.nz/iteration.html#unknown-sequence-length).
</div>

### Syntax

Let's simplify our repeated code by using a `for` loop. Below we'll discuss
the key components. Run the code below and verify the means match what we
saw in the last section.

```{r for_loop_syntax, exercise=TRUE}
output <- numeric(ncol(x))  # output vector to store results, line 1

for (i in seq_along(x)) {   # define our looping sequence, line 3
  output[i] <- get_column_mean(data = x, index = i) # code to iterate
}

output
```

### `for` loop components

1. Before the loop, you'll need to create an object to store your results. In
   the example above, we created a numeric vector that is the same length as
   the number of columns in our data frame `x`.
    ```{r include=FALSE}
    output <- numeric(ncol(x))
    output
    ```
    ```{r output_vec, exercise=TRUE}
    output <- numeric(ncol(x))
    output
    ```
   By default, each element of the vector is 0. Soon we'll fill these in with
   our mean calculations.
   
2. Next you'll need to define the looping sequence. This begins with the R
   keyword `for` and is wrapped in parentheses. The variable `i` is a dummy
   variable. Let's look at how `seq_along()` works with our data frame `x`.
   ```{r seq_along_example, exercise=TRUE}
   seq_along(x)
   ```
   This is a convenient way to get a vector from 1 to the number of columns in
   a data frame. The sequence, `i in seq_along(x)` means that each run
   of our loop, `i` will take on the next value in `seq_along(x)`, starting with
   1 the first time through the loop.
   
3. The last main component is the loop's body. This is code we have seen before.
   The code is run repeatedly each time for a different value of `i`. The
   first time through we would have
   ```{r include=FALSE}
   output[1] <- get_column_mean(data = x, index = 1)
   ```
   ```{r loop_body_1, exercise=TRUE}
   output[1] <- get_column_mean(data = x, index = 1)
    
   output
   ```
   The second time through the loop we would have
   ```{r loop_body_2, exercise=TRUE}
   output[2] <- get_column_mean(data = x, index = 2)
    
   output
   ``` 
   The iteration stops once `i` reaches the end of the looping sequence. In our
   small example, that would be once the code body is run for when `i` is 
   assigned the value of 4.
   
### Exercise 1

Object `mtcars` is a data frame
available in `base` R that comprises fuel consumption and 10 aspects of 
automobile design and performance for 32 automobiles in models from 1973 - 1974.
Below is a preview.

```{r preview_mtcars}
mtcars
```

Use a `for` loop to compute the median of every column in `mtcars`.

```{r ex1, exercise=TRUE}

```

```{r ex1-solution}
result <- numeric(length = ncol(mtcars)) # output vector

for (j in seq_along(mtcars)) {           # define looping sequence
  result[j] <- mtcars %>%                # loop body
    pull(j) %>% 
    median()
}

result
```


### Exercise 2

Use a `for` loop to compute the number of unique values in each column of the `planes` data frame
from the `nycflights13` package. Recall that this package has already been
loaded.
   
```{r ex2, exercise=TRUE}

```

```{r ex2-solution}
result <- numeric(length = ncol(planes))

for (i in 1:length(planes)) {
  result[i] <- planes %>% 
    pull(i) %>% 
    unique() %>% 
    length()
}

result
```

## Iteration caution

Many of the functions available in `base` R and other packages are vectorized.
This means that if you have a vector `x` with n elements and pass this
vector into a vectorized function, the function will be applied to each
element of `x`. For example:

```{r include=FALSE}
x <- -4:4
x
```

```{r create_x, exercise=TRUE}
x <- -4:4
x
```

```{r vectorize_ex_1, exercise=TRUE}
abs(x)
```

```{r vectorize_ex_2, exercise=TRUE}
sqrt(abs(x))
```

You could use a `for` loop to do the above computations, but this is not
recommended. It is inefficient, and you will not be harnessing R's full power
and functionality. Run the code below to verify you get the same answer
as above for taking the square root of the absolute value of `x`.

```{r bad_for_loop, exercise=TRUE}
x <- -4:4
output <- numeric(length(x))

for (k in 1:length(x)) {
  output[k] <- sqrt(abs(x[k]))
}

output
```

To determine if you'll need a loop to tackle your task or not, first look at 
the help for the functions you plan to work with to understand if they
are vectorized or not. Most functions you apply to a vector will be
vectorized and a loop will not be required. If you need to apply a function to
each column of a data frame or each component in a list, then you may need
to accomplish that task with a `for` loop.

## Check your understanding

Below are some questions to check your understanding on iteration with `for`
loops.

```{r quiz, echo=FALSE}
quiz(
    question("A `for` loop is primarily comprised of",
    answer("an output object, a looping sequence, and a body.",
           correct = TRUE,
           message = "These are the three main pieces."),
    answer("an output object, a looping sequence, a body, and a return statement.",
           message = ""),
    answer("an input object, a looping sequence, and a body.",
           message = ""),
    answer("an argument, a name, a looping sequence, and a body.",
           message = ""),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  
  question("`for` loops have a finite number of iterations.",
    answer("True", 
           correct = TRUE,
           message = "The loop ends once it reaches the end of the looping
                      sequence."),
    answer("False", 
           message = "The loop ends once it reaches the end of the looping
                      sequence."),
    random_answer_order = TRUE
  ),
  
  question("The `for` loop is the only type of loop that exists in R.",
    answer("True",
           message = "R also supports `while` and `repeat` loops."),
    answer("False", 
           correct = TRUE,
           message = "R also supports `while` and `repeat` loops."),
    random_answer_order = TRUE
  ),
  
  question("The R code `numeric(10)` will ",
    answer("create a numeric vector of length 10 of all zeros.",
           correct = TRUE,
           message = ""),
    answer("create a numeric vector of length 9 of all zeros.",
           message = ""),
    answer("create a numeric vector of the numbers 1 through 10.",
           message = ""),
    answer("create a numeric vector of the numbers 0 through 9.",
           message = ""),
    answer("create a null vector with enough memory for 10 elements.",
           message = ""),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  
  question("The dummy variable used in a `for` loop must be `i`.",
    answer("True",
           message = "Any variable name can be used. By convention, people
                      commonly use i, j, or k."),
    answer("False",
           correct = TRUE,
           message = "Any variable name can be used. By convention, people
                      commonly use i, j, or k."),
    random_answer_order = TRUE
  )
)
```

## References

1. Grolemund, G., & Wickham, H. (2020). 21 Iteration | R for Data Science. 
   Retrieved from https://r4ds.had.co.nz/iteration.html.
   
## Return to Learn R Tutorials Page

[Click here.](https://duke-learning-r.netlify.app)