---
title: "Customizing Data Visualizations"
author: "Maria Tackett"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r load-packages}
library(tidyverse)
library(patchwork)
library(flair)
```

In the [DATA VIZ - INSERT LINK] tutorial, you learned the basics of `ggplot2`, the tidyverse's primary package for creating data visualizations. In this tutorial, we will focus on different ways you can customize your data visualizations. 

By the end of this tutorial, you be able to:

- Add aesthetics to a plot
- Change the limits on the x and y axis
- Change the color palette
- Apply a theme to a plot
- Arrange multiple plots in a grid

## The Data

We will use the `mtcars` data set, which is automatically loaded when you open RStudio. The data is from the 1974 *Motor Trend* magazine, and it contains performance information and other specifications for 32 automobiles made in 1973 and 1974. 

To learn more about the data set and the variable definitions, type the following code in the console: 

```{r}
?mtcars
```

```{r}
mtcars <- mtcars %>%
  mutate(am = factor(am),
         cyl = factor(cyl), 
         vs = factor(vs)
         )
```

We create data visualizations using the `ggplot2` package. It is one of the core `tidyverse` packages, so it loads automatically when we load `tidyverse`.

*Notice `ggplot2` in the list of packages loaded in the code below.*

```{r, message = T, warning = T}
library(tidyverse)
```

## The basic plot 

Let's start with a scatterplot displaying the relationship between horsepower (`hp`) and miles per gallon (`mpg`).

```{r}
ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       title = "Horsepower vs. Miles per gallon")
```

Using this as a baseline, let's customize the plot. 


## Aesthetics 

We can add aesthetics (features) to our plot to incorporate additional variables or to customize the plot. These aesthetics include

- `color`
- `shape`
- `size`
- `alpha` (transparency)

Aesthetics can be mapped in two different ways: 

1. Aesthetic based on value of a variable. Aesthetic defined inside `aes()`.
2. Same aesthetic for all observations. Defined in the `geom_**` outside of `aes()`. 

Let's look at two plots in which the `color` aesthetic is applied in both ways: 

```{r echo = F}
p1 <- ggplot(data = mtcars, aes(x = hp, y = mpg, color = factor(am))) +
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon",
       color = "Transmission",
       subtitle = "by Transmission",
       title = "Horsepower vs. Miles per gallon")

p2 <- ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point(color = "blue") +
  labs(x = "Horsepower", 
       y = "Miles per gallon",
       title = "Horsepower vs. Miles per gallon")

p1 / p2 + plot_annotation(tag_levels = c('A'))
```

Notice in **Plot A** that the color of the points is based on the type of transmission (`am`). Using color in this way, we have added additional information to the plot, so instead of just doing a bivariate analsyis of the relationship between miles per gallon and horsepower, the visualization displays a multivariate relationship between horsepower, miles per gallon and transmission. 

Let's look at the code for **Plot A**. 

```{r aes-var, eval = F}
ggplot(data = mtcars, aes(x = hp, y = mpg, color = am)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Transmission",
       subtitle = "by Transmission",
       title = "Horsepower vs. Miles per gallon")
```

```{r echo = F}
decorate("aes-var") %>% 
  flair("color = am", background = "pink") %>%
  flair("subtitle = \"by Transmission\"", background = "pink") %>%
  flair("color = \"Transmission\"", background = "pink")
```

Notice how the code used to define the color is inside of the `aes()` function. This tells R to color the points based on the type of transmission (`am`). We also label the legend and add a subtitle to the plot. 

Now, let's take a look at the code for **Plot B**. 

```{r aes-all, echo = F, eval = F}
ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point(color = "blue") +
  labs(x = "Horsepower", 
       y = "Miles per gallon",
       title = "Horsepower vs. Miles per gallon")
```

```{r echo = F}
decorate("aes-all") %>% 
  flair("color = \"blue\"", background = "pink")
```

Here, we have applied the same aesthetic to all observations. Since we don't color the points based on the value of a variable, we define the color in the `geom_**` outside of the `aes()` function. In this case, we haven't added any additional information to the basic plot, but we have used the aesthetic to make the plot more visually appealing.

## Change axes limits

Let's stick with **Plot A** for now. 

Another way to customize the plot is to change the limits on the x and y axes. Suppose we want to zoom in on a portion of the plot. Let's change the axes to do so. 

```{r change-axis, echo = F, eval = F}
ggplot(data = mtcars, aes(x = hp, y = mpg, color = am)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Transmission",
       subtitle = "by Transmission",
       title = "Horsepower vs. Miles per gallon") +
  xlim(c(100, 300)) +
  ylim(c(15, 25))
```

```{r echo = F}
decorate("change-axis") %>% 
  flair("xlim(c(100, 300)) +", background = "pink") %>%
  flair("ylim(c(15, 25))", background = "pink")
```

## Change scale / color palette - come back to this!

We can change the color palette used for the visualization. The default color palette used in `ggplot2` is not color-blind safe, so we can change the color palette to make our plots more accessible to all readers. 

Let's look at a plot of the Miles per gallon vs. Horsepower colored by the cylinders.

```{r}
ggplot(data = mtcars, aes(x = hp, y = mpg, color = cyl)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Cylinder",
       subtitle = "by Cylinder",
       title = "Horsepower vs. Miles per gallon") 
```

This is the default color palette in `ggplot2`; however,we may want to use a different color palette. One way to do this is by using the `scale_color_brewer()` function. Using this function, we can indicate which RColorBrewer palette we wish to use. 

Here is the same plot above using the "Set1" color palette: 

```{r set1-color, eval = F, echo = F}
ggplot(data = mtcars, aes(x = hp, y = mpg, color = cyl)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Cylinder",
       title = "Horsepower vs. Miles per gallon",
       subtitle = "by Cylinder") + 
  scale_colour_brewer(palette = "Set1")
```

```{r echo = F}
decorate("set1-color") %>%
  flair("scale_colour_brewer(palette = \"Set1\")", background = "pink")
```

Another option for color palettes are those available in the `viridis` package. These palettes are good to make your plots more readable by those with colorblindness. 

```{r viridis, echo = F, eval = F}
library(viridis)  

ggplot(data = mtcars, aes(x = hp, y = mpg, color = cyl)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Cylinder",
       title = "Horsepower vs. Miles per gallon",
       subtitle = "by Cylinder") + 
   scale_color_viridis(discrete = TRUE) # for discrete color scale
```
```{r echo = F}
decorate("viridis") %>%
  flair("library(viridis)", background = "pink") %>%
  flair("scale_color_viridis(discrete = TRUE) # for discrete color scale", background = "pink")
```


You can find a full list of color palettes to use in `scale_color_brewer` and `scale_color_viridis` [here](https://www.datanovia.com/en/blog/top-r-color-palettes-to-know-for-great-data-visualization/).


## Change theme

We may want to change the overall look of the visualization, e.g. the background, gridlines, etc. One way to do this is by applying a theme to the visualization. 

### Themes in `ggplot2` package

Let's look at two different themes that are available in `ggplot2` package.

```{r theme-dark, echo = F, eval = F}
ggplot(data = mtcars, aes(x = hp, y = mpg, color = am)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Transmission",
       subtitle = "by Transmission",
       title = "Horsepower vs. Miles per gallon") +
 theme_dark()
```

```{r echo = F}
decorate("theme-dark") %>%
  flair("theme_dark()", background = "pink")
```

```{r theme-classic, eval = F, echo = F}
ggplot(data = mtcars, aes(x = hp, y = mpg, color = am)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Transmission",
       subtitle = "by Transmission",
       title = "Horsepower vs. Miles per gallon") +
 theme_classic()
```

```{r echo = F}
decorate("theme-classic") %>%
  flair("theme_classic()", background = "pink")
```

A list of complete themes available in the `ggplot2` package can be found [here](https://ggplot2.tidyverse.org/reference/ggtheme.html). 

You can also customize the elements of a visualization individually by using the `theme()` function. One instance in which this function is especially useful is to update the font size for the labels on plots. 

```{r font-size, echo = F, eval = F}
ggplot(data = mtcars, aes(x = hp, y = mpg, color = am)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Transmission",
       subtitle = "by Transmission",
       title = "Horsepower vs. Miles per gallon") +
 theme(plot.title = element_text(size = 20), 
       axis.title.x = element_text(size = 16), 
       axis.title.y = element_text(size = 16))

```

```{r echo = F}
decorate("font-size") %>%
  flair("theme(plot.title = element_text(size = 20),", background = "pink") %>%
  flair("axis.title.x = element_text(size = 16),", background = "pink") %>%
  flair("axis.title.y = element_text(size = 16))", background = "pink")
```

See the [`theme()` reference page](https://ggplot2.tidyverse.org/reference/theme.html) to learn more about customizing themes. 

### `ggthemes` package

The `ggthemes` package has a selection of custom themes and color palettes to help you customize your plot. You can find all of the themes and palettes on the [`ggthemes` webpage](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/). 

You will need to load the `ggthemes` package. Here is an example of one of their themes: 

```{r ggthemes, eval = F, echo = F}
library(ggthemes)

ggplot(data = mtcars, aes(x = hp, y = mpg, color = am)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Transmission",
       subtitle = "by Transmission",
       title = "Horsepower vs. Miles per gallon") +
 theme_fivethirtyeight()
```

```{r echo = F}
decorate("ggthemes") %>%
  flair("library(ggthemes)", background = "pink") %>%
  flair("theme_fivethirtyeight()", background = "pink")
```


### Patchwork 

Sometimes we want to organize multiple plots in a grid to more neatly format our document / presentation. This is especially helpful if you have a lot of plots (e.g. when doing exploratory data analysis). 

We can use the `patchwork` package to more neatly organize plots in a grid. To do so, you can assign an object name to each plot, then use those names to organize the plots. 

Let's suppose we have the following three plots assigned to names `p1`, `p2`, and `p3`: 

```{r}
p1 <- ggplot(data = mtcars, aes(x = hp, y = mpg, color = am)) + 
  geom_point() +
  labs(x = "Horsepower", 
       y = "Miles per gallon", 
       color = "Transmission")

p2 <- ggplot(data = mtcars, aes(x = cyl)) + 
  geom_bar() +
  labs(x = "Number of Cylinders")

p3 <- ggplot(data = mtcars, aes(x = hp)) + 
  geom_histogram() +
  labs(x = "Horsepower")
```

First, let's use the code below to load the patchwork package: 

```{r echo = F}
library(patchwork)
```

Use `/` to put a plot on a new row:  

```{r}
p1 / p2 / p3
```

Use `+` to put a plot in a new column: 

```{r}
p1 + p2 + p3
```

You can combine the `/` and `+` to arrange the plots by row and column. Let's put `p1` and `p2` on the first row and `p3` on the second row: 

```{r}
(p1 + p2) / p3
```
Just like in "order of operations", the `/` will be applied before `+`, so use parentheses (as in the code above) if you want the `+` to apply first. 

Let's repeat the code above without the parentheses: 

```{r}
p1 + p2 / p3
```

Notice how the plots differ. 

Finally, you can use the `plot_annotation()` function to add labels and other text annotations to patchwork objects. Using the code below, we can use letters to label the plots: 

```{r}
(p1 + p2) / p3 +
  plot_annotation(tag_levels = "A")
```

## Practice: Put it all together





## More resources

- https://ggplot2.tidyverse.org/
- https://patchwork.data-imaginist.com/index.html
- [R for Data Science: Data visualization](https://r4ds.had.co.nz/data-visualisation.html)
- [ggplot2: Elegant Graphics for Data Analysis](ggplot2: Elegant Graphics for Data Analysis)

