---
title: "Basics of ggplot"
output: 
  learnr::tutorial:
    progressive: false
    theme: "cerulean"
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome to ggplot2

The package `ggplot2` is a tidyverse package that allows you to make beautiful graphics in R! This tutorial will cover how you can make various graphics. You will learn commands that you can use to customize your plots. This tutorial builds on the discussion of `ggplot2` in the tidyverse tutorial, so if you've seen that one, you will recognize the design of the first part of this tutorial, although it makes use of different data. For more specific information on the aesthetics of plots, please check out the tutorial on customizing plots.

## Basics of ggplot2

`Ggplot2` is one of the most used tidyverse packages to visualize data. This tutorial discusses the basics of `ggplot2` and then presents specific plots and how to customize them.

`Ggplot2` is based on idea of having a [grammar of graphics](https://cfss.uchicago.edu/notes/grammar-of-graphics/) where we build plots by layering objects on top of one another. This allows us to start from the base and build up, having control of each feature of the plot.

The basic command that you begin with for `ggplot2` is `ggplot()`. Inside the parentheses, you then tell ggplot what data you are using and what the x and y variables are for that visualization. Inside the parentheses of `ggplot()` you will first write the name of the data set, then put a comma, and then to identify the x and y variables write `aes(x = , y = )`. In some cases, you will just be looking at counts of data, so you will just have an x= here and no y.

In this tutorial, I use data from states that held Democratic primaries on [Super Tuesday](https://www.washingtonpost.com/politics/2020/02/19/what-is-super-tuesday/?arc404=true) this year. The election data I use here comes from the [New York Times](https://www.nytimes.com/interactive/2020/03/03/us/elections/results-super-tuesday-primary-election.html) and the exit poll results come from [CNN](https://www.cnn.com/election/2020/entrance-and-exit-polls/). (States/territories where there was not an exit poll are omitted.) This data is available to you separately and you should change the file path so that you can load it into R on your computer if you want to replicate these graphics.

First, I make a histogram of former Vice President Joe Biden's vote percentage by state using the hist command. Let's say that you wanted to make a plot using `ggplot()` and "Location" was your x variable and "Percentage" was your y variable. Your code would look like this:

```{r basics, echo=TRUE, fig.width=5}
library(readxl)
library(ggplot2)
# You could also load all of the tidyverse packages here with library(tidyverse)
SuperTuesday <- read_excel("D:/USB20FD - Copy/SuperTuesday.xlsx")
ggplot(SuperTuesday, aes(x = State, y = Pct_Biden))
```

This has set up the foundation for our plot, but hasn't actually put any visualization on the grid. To do that, you need to add something else to your code. Let's say that we want a bar chart that shows the percentage of the vote that Biden received in each state. You would add ` + geombar(stat = identity)` to your code. Overall, your code would now look like this and now your plot isn't just a blank grid.

```{r barplot, echo=TRUE, fig.width=5}
ggplot(SuperTuesday, aes(x = State, y = Pct_Biden)) +
  geom_bar(stat = "identity")
```

## Histograms in ggplot2
Bar charts are one nice way to visualize this data, but you might also be interested in grouping states together by Biden vote percentage to see the typical values for this variable. To do that, you no longer would have a y variable because you want the y axis to represent counts of the number of states that fall into that bin (e.g., between 20 and 25% of the vote). You can also set the bin size in R by using the `binwidth=` command to tell R how wide to make the bins. The initial binwidth is set fairly narrowly, so you will often want to use this command. You can see here that Biden got around 35 to 40 percent in the largest number of states.

```{r histogram, echo=TRUE, fig.width=5}
ggplot(SuperTuesday, aes(x = Pct_Biden)) +
  geom_histogram(binwidth = 5)
```

## Customizing the Colors on Your Plot
Once you've created a plot in `ggplot2`, there are a bunch of other neat things you can do with it, including changing the color of both the interior and border of the bars. See here how you could make your bars green and outlined in orange.

Here I use hex codes that draw from a [color-blind friendly palette](https://www.color-hex.com/color-palette/49436), but you can also use names of colors in `ggplot`.
```{r histogram_colors, echo=TRUE, fig.width=5}
ggplot(SuperTuesday, aes(x = Pct_Biden)) +
  geom_histogram(binwidth = 5, color = "#d55e00", fill = "#009e73")
```

## Customize Your Axis Labels

Notice on this plot that the label for the x-axis is "Pct_Biden" and the label on the y-axis is "counts" and there is no overall title. If you want to change or add axis labels or an overall plot title, use the commands `xlab`, `ylab`, and `ggtitle`, respectively, for the x-axis, y-axis, and overall plot. Notice that R automatically aligns the main title left, so I added a line of code to align the code over the center of the histogram using the `theme()` command. To align center, use `hjust(0.5)`. If you would prefer to align it further to the left, use a number below 0.5, while a number greater than 0.5 will push it right.

```{r histogram_labels, echo=TRUE, fig.width=5}
ggplot(SuperTuesday, aes(x = Pct_Biden)) +
  geom_histogram(binwidth = 5, color = "#d55e00", fill = "#009e73") +
  xlab("Percentage of Vote") +
  ylab("Number of States") +
  ggtitle("Biden Vote Percentage on Super Tuesday") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Making Subplots
Now you know how to make a histogram of the entire data set. But what if you wanted to have a multi-panel plot with data separated by region on separate graphs? To do that, you could use `facet_wrap()` or `facet_grid()`. Here, let's imagine that we wanted to make a different plot for the North and the South. By adding `facet_grid(~South)` we can divide the data into two panels by column, one for states in the North and another for states in the South. (If you wanted to divide into a rectangular layout, you could instead use `facet_wrap()`.) Clearly, Biden tended to do better in the South than in the North. 

```{r facet, echo=TRUE, fig.width=5}
ggplot(SuperTuesday, aes(x = Pct_Biden)) +
  geom_histogram(binwidth = 5, color = "#d55e00", fill = "#009e73") +
  xlab("Percentage of Vote") +
  ylab("Number of States") +
  ggtitle("Biden Vote Percentage on Super Tuesday") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_grid(~South)
```

We only have a small number of data points here, but if we wanted to divide the panels by a second variable in addition to South, we would add that to `facet_grid` by adding " + variable name" inside the parentheses. For example, if we had a variable for state winner, our expression would now be `facet_grid(~South + winner)`.

## Scatterplot: Comparing Two Variables

Sometimes, you might like to compare two variables at once. Let's say that you thought that there might be an association between Biden's percentage of the vote in a state and the percentage of the Democratic electorate that was Black. To make a scatter plot, you can use the `geom_point()` command. If you would like to label the individual points so that the person reading your report knows which state is which, you can add these labels by specifying the name of the variable you would like to use as the label inside `aes()` and by then adding `geom_text()` later. Here is an example of how to make a scatter plot. You can add ``geom_smooth(method = 'lm', se = FALSE) `` to add a least squares line.

```{r scatterplot, echo=TRUE, fig.width=5}
ggplot(SuperTuesday, aes(x = Pct_Black, y = Pct_Biden, label = State)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Percent Black") +
  ylab("Percent Biden") +
  ggtitle("Does Biden Do Better in States with More Black Voters?") +
  geom_text() +
  theme(plot.title = element_text(hjust = 0.5))
```

## Visualizing Time Series Data in ggplot2

So far, the data we have looked at comes from a single point in time (i.e., it is cross-sectional data), Sometimes, you will want to plot a variable (or several) over an extended period of time. Here, as an example, I use data from US House elections since the end of World War II (i.e., elections from 1946 to 2018) The below line plot displays the number of House seats won by the president's party in each election.

```{r house2, echo=TRUE, fig.width=5}
House <- read_excel("D:/USB20FD - Copy/House.xlsx")
ggplot(House, aes(x = ElectionYear, y = SeatsWon)) +
  geom_line() +
  xlab("Year") +
  ylab("Number of Seats Won") +
  ggtitle("Seats in the House for the President's Party") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Adding Multiple Lines and Making a Legend

Sometimes you might want to have multiple lines on your line plot. Here, I create lines for the number of seats won by the president's party before the election and for the number of seats held by the party before the election. I make the number of seats won in that year's election steel blue and the number of seats held before the election in pink.

Another thing you may want to do is add a legend. To do this, I used ```scale_color_manual```. Before doing this, you specify the name for that variable that you want to appear in the legend in ```aes```. Then, using ```scale_color_manual```, you can pick a color for each variable, using the name you picked in ```aes```.

```{r house3, echo=TRUE, fig.width=5}
ggplot() +
  geom_line(data = House, aes(y = SeatsWon, x = ElectionYear, color = "Seats Won"), size = 1) +
  geom_line(data = House, aes(y = SeatsHeld, x = ElectionYear, color = "Seats Held"), size = 1) +
  scale_color_manual(name = "Number of Seats", values = c(
    "Seats Won" = "steelblue",
    "Seats Held" = "pink"
  )) +
  labs("Number of Seats") +
  xlab("Election Year") +
  ylab("Number of Seats") +
  ggtitle("Seats in the House for the President's Party") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Changing Tick Mark Labels
RStudio gives us three tick marks and labels on the x-axis at 1960, 1980, and 2000. What if we wanted more (or fewer) here? With a continuous variable such as year, you would add scale_x_continuous to your code and then specify the labels you would like to have on the x-axis. If you wanted to change the y-axis you would use `scale_y_continuous` instead. If you were using a different kind of plot with a discrete x- or y-axis variable, you would use `scale_x_discrete` or `scale_y_discrete`, respectively. An example of how to apply this code to the election data plot by adding a tick mark for the start of each decade is below.

```{r house, echo=TRUE, fig.width=5}
ggplot() +
  geom_line(data = House, aes(y = SeatsWon, x = ElectionYear, color = "darkblue"), size = 1) +
  geom_line(data = House, aes(y = SeatsHeld, x = ElectionYear, color = "red"), size = 1) +
  scale_x_continuous(breaks = c(1950, 1960, 1970, 1980, 1990, 2000, 2010)) +
  scale_color_discrete(name = "Number of Seats", labels = c("Seats Won", "Seats Held")) +
  xlab("Election Year") +
  ylab("Number of Seats") +
  ggtitle("Seats in the House for the President's Party") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Other Visualizations in ggplot2
There are many other visualizations you can make in `ggplot2`. Each of them has a specific command that goes with them in the code. For example, for a density plot, you use `geom_density` and for a dot plot you use `geom_dotplot`. [This](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) cheat sheet provides a good guide to specific types of plots you can make, but the same broad pattern exists across different visualizations.

## Quiz
Here are two quick quiz questions to review the material in this tutorial:

```{r quiz}
quiz(
  question("If you wanted to make a histogram with bins of width 10, how you would do it inside geom_histogram()?",
    answer("bin: 10"),
    answer("bins = 10"),
    answer("binwidth = 10", correct = TRUE),
    answer("bin_width = 10"),
    allow_retry = TRUE
  ),
  question("Which of these allow you to load `ggplot2` into R? Please select all that apply.",
    answer("library(ggplot2)", correct = TRUE),
    answer("library(ggplot)"),
    answer("library(plots)"),
    answer("library(tidyverse)", correct = TRUE),
    allow_retry = TRUE
  )
)
```
