---
title: "Bootstrapping with Infer"
output: 
  learnr::tutorial:
    progressive: false
    theme: "cerulean"
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  exercise.timelimit = 3
)
```

## Introduction

---

This brief tutorial will help you understand bootstrapping and learn how to 
construct bootstrap confidence intervals using the `infer` package. 
To start, load the `tidyverse` and `infer` packages. If you haven't already 
installed `infer`, do that first by running the code below.

```{r eval = FALSE}
install.packages("infer")
```

Remember, you only need to install packages once, but you need to load them at 
the start of every session.

```{r load-packages}
library(tidyverse)
library(infer)
```

Recall that statistical inference refers to the process of generalizing
from a sample to a population. We could use a single number (a 
point estimate) as our best guess of the population parameter but we often 
prefer a range of plausible values (an interval estimate or a confidence 
interval).

To construct a confidence interval we need to quantify the variability of 
our sample statistic. We can do this using the Central Limit Theorem but
today we will focus on a simulation-based method called bootstrapping. Invented 
in 1979 by Brad Efron, bootstrapping allows us to construct confidence intervals
based on our sample data using resampling.

In general, bootstrap methods

- have fewer requirements than other (CLT-based) methods
- give more accurate answers
- are more flexible and easily applied to new settings

The basic idea is as follows: our sample is representative of 
the population, so if we resample from our original sample many times, this is 
representative of what we would obtain if we took many samples 
from the population.

To bootstrap we follow the steps outlined below:

1. Take a bootstrap sample: randomly sample **with replacement** from the 
original sample using the same sample size as the original sample.
2. Find the bootstrap statistic: calculate the statistic of interest using 
the resampled data. This can be the mean, median, proportion, slope, etc.
3. Create the bootstrap distribution: Repeat steps (1) and (2) many times using 
a computer.
4. Construct a confidence interval using the bootstrap distribution.

That's it!

## Exploring the Data

To introduce the bootstrap idea, we will use data and material from Tim 
Hesterberg's Bootstrap Methods and Permutation Tests. A link is provided in the 
Resources and References panel below.

The telecommunications conglomerate Verizon is both a local telephone company 
that maintains land-line phone services, called an Incumbent Local Exchange 
Carrier (ILEC), and a Competitive Local Exchance Carrier (CLEC), which means it
sells long-distance services. If there is a service outage and a repair is 
needed, Verizon is obligated by law to make repairs as quickly for CLEC 
customers as their own. We will investigate the repair times (in hours) for the
two types of customers.

The data is in the `Verizon.csv` file. We read in the data and create subsets 
containing the ILEC and CLEC repair times.

```{r}
verizon <- read_csv("data/verizon.csv")
verizon_ilec <- verizon %>% filter(Group == "ILEC")
verizon_clec <- verizon %>% filter(Group == "CLEC")
```

Create an appropriate plot of the repair times for ILEC and CLEC customers.

```{r ex1, exercise=TRUE}

```

```{r ex1-solution}
ggplot(verizon, aes(x = Group, y = Time)) +
  geom_boxplot()
```

Calculate summary statistics (including mean, median, sd, and n) for the 
repair times for ILEC and CLEC customers.


```{r ex2, exercise=TRUE}

```

```{r ex2-solution}
verizon %>%
  group_by(Group) %>%
  summarize(
    mean_wait = mean(Time),
    median_wait = median(Time),
    sd_wait = sd(Time),
    count = n()
  )
```

You should notice that the mean time is higher for CLEC customers than for ILEC 
customers and that the data are non-normal. In particular, the ILEC repair 
times are extremely right-skewed.

## Bootstrapping with Infer

To generate the bootstrap distribution using the infer package we must:

- `specify()` the variable of interest
- `generate()` the bootstrap samples
- `calculate()` the bootstrap statistics

Let's use the three functions above to construct a bootstrap confidence 
interval for the mean repair time for CLEC customers. We use the `set.seed()` 
function below to control the random number generator and make our simulation 
reproducible.

```{r}
set.seed(42)

boot_dist <- verizon_clec %>%
  specify(response = Time) %>%
  generate(reps = 10000, type = "bootstrap") %>%
  calculate(stat = "mean")
```

We `specify` that the `Time` column in the `verizon_clec` data is our 
response of interest, `generate` 10000 resamples from our original sample of 
type bootstrap, and `calculate` the mean for each sample. These three functions 
generate the bootstrap distribution and store the results in `boot_dist`.


```{r quiz, echo=FALSE}
quiz(
  question("How many rows are in `boot_dist`?",
    answer("10,000",
      correct = TRUE,
      message = "We specified this using the generate function."
    ),
    answer("23",
      message = ""
    ),
    answer("1664",
      message = ""
    ),
    answer("2",
      message = ""
    ),
    answer("1000",
      message = ""
    ),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),

  question("What is in each row of `boot_dist`?",
    answer("Each row in `boot_dist` is a bootstrapped mean - the mean of 
           one bootstrapped sample.",
      correct = TRUE,
      message = ""
    ),
    answer("Each row in `boot_dist` is our original sample mean.",
      message = ""
    ),
    answer("Each row in `boot_dist` is a bootstrap sample - a sample taken with
           replacement from our original sample.",
      message = ""
    ),
    random_answer_order = TRUE
  ),
  question("For each bootstrap sample, how many draws with replacement were 
               taken from the original sample?",
    answer("10,000",
      message = ""
    ),
    answer("23",
      correct = TRUE,
      message = "This is the CLEC sample size."
    ),
    answer("1664",
      message = ""
    ),
    answer("2",
      message = ""
    ),
    answer("1000",
      message = ""
    ),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

Let's visualize the bootstrap distribution. We can do this using `ggplot`.

```{r}
ggplot(data = boot_dist, aes(x = stat)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Bootstrapped Mean Times", y = "Count")
```

Alternatively, the `infer` package has a built-in function called `visualize` 
that allows us to visualize the bootstrap distribution.

```{r}
visualize(boot_dist) +
  labs(x = "Bootstrapped Mean Times", y = "Count")
```

To construct a 95\% bootstrap confidence interval, just find the middle 95\% of
the bootstrap distribution. The middle 95\% corresponds to the lowest 2.5\% of 
the data and the top 97.5\% of the data, so we use the `quantile` function with
values 0.025 and 0.975. We can use the same procedure for any desired level of 
confidence.


```{r}
boot_dist %>%
  summarize(
    lower_bound = quantile(stat, 0.025),
    upper_bound = quantile(stat, 0.975)
  )
```

The `infer` package also has a built-in function called `get_ci` which does the 
same thing.

```{r}
percentile_ci <- get_ci(boot_dist, level = 0.95)
percentile_ci
```

Finally, let's visualize the confidence interval on the bootstrap distribution.

```{r}
lower_bound <- boot_dist %>%
  summarize(lower_bound = quantile(stat, 0.025)) %>%
  pull()
upper_bound <- boot_dist %>%
  summarize(upper_bound = quantile(stat, 0.975)) %>%
  pull()

ggplot(data = boot_dist, aes(x = stat)) +
  geom_histogram(binwidth = 1) +
  geom_vline(xintercept = c(lower_bound, upper_bound), linetype = "dotted") +
  labs(x = "Bootstrapped Mean Times", y = "Count")
```

We can accomplish the same thing using tools from the `infer` package. Remember, 
if you are ever confused about a function you can pull up the documentation by
running (for example) `?visualize`.

```{r}
visualize(boot_dist) +
  shade_confidence_interval(endpoints = percentile_ci) +
  labs(
    x = "Bootstrapped Mean Times", y = "Count",
    caption = "Green lines represent 95% C.I. bounds"
  )
```

Finally, we provide a brief one-sentence interpretation in the context of the 
problem. We are 95\% confident that the true mean repair time for CLEC repairs 
is between 10.2 and 25.7 hours.

## Your turn!

1. Fill in the blanks in the code chunk below to generate the bootstrap
distribution of the median repair time for ILEC customers.

```{r ex3, exercise=TRUE}
set.seed(10)
```

```{r ex3-solution}
set.seed(10)
boot_dist <- verizon_ilec %>%
  specify(response = Time) %>%
  generate(reps = 10000, type = "bootstrap") %>%
  calculate(stat = "median")
```


2. Fill in the blanks in the code chunk below to construct an 80\% bootstrap 
confidence interval using the bootstrap distribution in exercise #1. Provide 
an interpretation of your interval in context.

```{r ex4, exercise=TRUE}

```

```{r ex4-solution}
get_ci(boot_dist, level = 0.80)
```

3. Construct a 90% bootstrap confidence interval using the ILEC repait time 
data using a statistic of your choosing other than the median or mean and
provide a one-sentence interpretation of your interval. 

```{r ex5, exercise=TRUE}

```




## Resources and References

- [Infer Website](https://infer.netlify.app/)
- [Computer Intensive Methods in Physics by Persi Diaconis and Bradley Efron](
https://statistics.sites.stanford.edu/sites/g/files/sbiybj6031/f/BIO%2083.pdf)
- [Statistical Inference using Bootstrap Confidence Intervals by Michael Wood](
https://rss.onlinelibrary.wiley.com/doi/pdf/10.1111/j.1740-9713.2004.00067.x)
- [Bootstrap Methods and Permutation Tests by Tim Hesterberg](
https://statweb.stanford.edu/~tibs/stat315a/Supplements/bootstrap.pdf)
