---
title: "CLT-Based Inference with Infer"
output:
  learnr::tutorial:
    progressive: no
    theme: cerulean
    allow_skip: yes
runtime: shiny_prerendered
resource_files:
- clt_inference.Rmd
- clt_inference.Rmd
- clt_inference.Rmd
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
library(tidyverse)
library(infer)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  exercise.timelimit = 3
)
```

```{r include=FALSE}
library(learnr)
library(knitr)
```

## Introduction

In this tutorial you will learn the basics of CLT-based inference using the 
`infer` package. To begin we load the necessary packages. Remember, if you 
haven't already installed `infer`, do that first.

```{r eval = FALSE}
install.package("infer")
```

```{r load-packages, eval = FALSE}
library(tidyverse)
library(infer)
```

Recall the **Central Limit Theorem**. For a population with mean $\mu$ and 
standard deviation $\sigma$ the distribution of $\bar{X}$ is normal, centered 
at $\mu$, and with variability inversely proportional to the square root of the
sample size.

$$Z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} \sim N(0,1)$$ 

We often don't know $\sigma$ so we estimate from our sample data using $s$. The 
test statistic below has a $t$-distribution with $n-1$ degrees of freedom.

$$T = \frac{\bar{X} - \mu}{s/\sqrt{n}} \sim t_{n-1}$$ 

Relative to the normal distribution, the $t$ distribution has thicker tails to
make up for the additional variability incurred by using $s$ to estimate 
$\sigma$.

## Exploring the Data

We will use the `gss` data from the `infer` package. This data set contains a 
sample of 500 entries from the General Social Survey (GSS). The GSS is a 
national survey of American society and opinions that has been conducted since
1972.

First, familiarize yourself with the data set and variables using the code 
below. You can learn more about the General Social Survey by visiting their 
[website](https://gss.norc.org/).

```{r eval = FALSE}
?gss
```

We will focus on the hours variable, the number of hours worked in the week 
before the survey, truncated at 89. We are interested in determining if the 
sample data provide convincing evidence that Americans, on average, work more 
than 40 hours per week.

Let's first conduct a brief exploratory data analysis. Begin by creating a 
well-labeled histogram of the number of hours worked per week.

```{r ex1, exercise = TRUE}

```

```{r ex1-solution}
ggplot(data = gss, aes(x = hours)) +
  geom_histogram(binwidth = 2) +
  labs(x = "Hours Worked Per Week", y = "Count")
```

Next, calculate summary statistics for the number of hours worked per week, 
including the sample mean `xbar`, standard deviation `s`, and sample size `n`.
Store the results in `hours_summary`.

```{r ex2, exercise = TRUE}

```

```{r ex2-solution}
hours_summary <- gss %>%
  summarize(
    xbar = mean(hours),
    s = sd(hours),
    n = n()
  )
hours_summary
```


```{r echo = FALSE}
hours_summary <- gss %>%
  summarize(
    xbar = mean(hours),
    s = sd(hours),
    n = n()
  )
```

## Hypothesis Testing

Does the sample data provide convincing evidence that Americans, on average, 
work more than 40 hours per week?

We first write out the appropriate null and alternative hypothesis. 

- $H_o: \mu = 40$
- $H_a: \mu > 40$

We will calculate the test statistic step-by-step. First pull our summary data 
from `hours_summary`.

```{r}
n <- hours_summary %>% pull(n)
s <- hours_summary %>% pull(s)
xbar <- hours_summary %>% pull(xbar)
```

Next, we calculate the standard error using `s` and `n`.

```{r}
se <- s / sqrt(n)
```

Then, we use the sample mean `xbar` and the standard error `se` to calculate the
test statistic.

```{r}
t <- (xbar - 40) / se
```

The test statistic is about `r round(t, 2)`. 

```{r}
df <- n - 1
```

And the degrees of freedom is `r df`.

Finally, we use the test statistic `t` and the `pt()` function in R to find the
$p$-value.

```{r}
pt(t, df, lower.tail = FALSE)
```

The $p$-value is small, so we reject the null hypothesis. These data provide
sufficient evidence at $\alpha = 0.05$ to conclude that Americans, on average,
work more than 40 hours per week.

We can use the `visualize` function from the `infer` package to plot the 
CLT-based null distribution. We indicate the test statistic with a red 
vertical line and shade the region corresponding to the $p$-value in red.

```{r}
clt_based_null_distribution <- gss %>%
  specify(response = hours) %>%
  hypothesize(null = "point", mu = 40) %>%
  calculate(stat = "t")

clt_based_null_distribution %>%
  visualize(method = "theoretical") +
  shade_p_value(t, direction = "greater")
```

## Confidence Intervals

Next we construct a confidence interval for the mean. Do you expect the 
confidence interval to include 40?

```{r}
t_star <- qt(.95, df)
xbar + c(-1, 1) * t_star * se
```

We are 95\% confident that the true average hours worked per week for adult 
Americans is between `r round(xbar + -1 * t_star * se, 2)` and 
`r round(xbar + 1 * t_star * se, 2)`.

Instead of coding each step individually every time we want to do a hypothesis 
test, we can use the `t_test` function from the `infer` package to conduct a
$t$-test. To do this we need to specify the following:

- `x`: a tibble
- `response`: the variable in `x` that serves as the response
- `mu`: the hypothesized null mean value
- `conf_int`: a logical value indicating whether to include a confidence 
interval
- `conf_level`: the confidence level

```{r}
t_test(
  x = gss,
  response = hours,
  mu = 40,
  conf_int = FALSE,
  alternative = "greater",
  conf_level = 0.95
)
```

We reject the null hypothesis. These data provide sufficient evidence using 
$\alpha = 0.05$ to conclude that Americans, on average, work more than 40 hours 
per week.

## Your turn!

The GSS asks respondents to report their age at the time of the survey. Do these 
data provide convincing evidence that the average age of American adults is 
greater than 30?

- $H_o: \mu = 30$
- $H_a: \mu > 30$

1. Calculate summary statistics and store the sample size as `n`, the standard 
deviation as `s` and the sample mean as `xbar`.

2. Use the summary statistics and standard error to calculate the test 
statistic. 

3. Calculate the $p$-value.

```{r ex3, exercise = TRUE}

```

```{r ex3-solution}
age_summary <- gss %>%
  summarize(
    xbar = mean(age),
    s = sd(age),
    n = n()
  )

n <- age_summary %>% pull(n)
s <- age_summary %>% pull(s)
xbar <- age_summary %>% pull(xbar)

se <- s/sqrt(n)
t <- (xbar - 30) / se
t

df <- n - 1
pt(t, df, lower.tail = FALSE)
```

```{r quiz, echo=FALSE}
quiz(
  question("What is the appropriate statistical decision? Include a conclusion 
           in the context of the problem.",
    answer("We reject the null hypothesis.",
      correct = TRUE,
      message = "At a significance level of 5%, the data provide 
           convincing evidence that the average age of American adults is
           greater than 30."
    ),
    answer("We fail to reject the null hypothesis.",
      message = ""
    ),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

4. Use the code chunk below to replicate the analysis above using the
`t_test` function from the `infer` package.

```{r ex4, exercise = TRUE}

```

```{r ex4-solution}
t_test(
  x = gss,
  response = age,
  mu = 30, alternative = "greater"
)
```

5. In additional to hours worked per week, the GSS asked if the respondent 
received a college degree. Use the `t_test` function from the `infer` package 
to test if Americans with a college degree work the same number of hours per 
week as Americans without a college degree. Remember, if you are confused 
about a function you can pull up the documentation using `?t_test`.

```{r ex5, exercise = TRUE}

```

```{r ex5-solution}
t_test(
  x = gss,
  formula = hours ~ college,
  order = c("degree", "no degree"),
  alternative = "two-sided"
)
```

```{r quiz2, echo=FALSE}
quiz(
  question("Report the p-value from Exercise #5 and provide an interpretation 
  of the p-value in the context of the problem.",
    answer("0.264",
      correct = TRUE,
      message = "If there was no relationship at all between college degree 
      and hours worked per week, the probability that we would see a test 
      statistic as or more favorable to the alternative hypothesis as what 
      we observed is 0.264."
    ),
    answer("0.05",
      message = ""
    ),
    answer("1.12",
      message = ""
    ),
    answer("0.112",
      message = ""
    ),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Resources and References

- [Infer Website](https://infer.netlify.app/)
- [General Social Survey Website](https://gss.norc.org/)

## Return to Learn R Tutorials Page

[Click here.](https://duke-learning-r.netlify.app)
