---
title: "Simulation-Based Inference with Infer"
output:
  learnr::tutorial:
    progressive: no
    theme: cerulean
    allow_skip: yes
runtime: shiny_prerendered
resource_files:
- sim_inference.Rmd
---

## Introduction

```{r setup, include=FALSE}
library(tidyverse)
library(infer)

cheating_data <-
  tibble(outcome = c(rep("text", 27), rep("no text", 267 - 27)))
phat <- gss %>%
  specify(response = sex, success = "female") %>%
  calculate(stat = "prop") %>%
  pull(stat)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r include=FALSE}
library(learnr)
library(knitr)
```

In this tutorial you will learn the basics of simulation-based inference using
the `infer` package. To begin we load the necessary packages. Remember, if you 
haven't already installed `infer`, do that first.

```{r load-packages}
library(tidyverse)
library(infer)
```

Recall statistical hypothesis testing is the procedure that allows us to decide
between two claims about the population parameter. We will follow a four step
procedure to conduct a hypothesis.

1. State the null and alternative hypothesis.

2. Take a representative sample of the population, collect data, and summarize 
the data with a statistic.

3. Determine how likely it is to see data as or more surprising as what we 
observed, assuming the null hypothesis is true.

4. Make a decision. If our data is extremely unlikely assuming the null 
hypothesis is true, we reject the null hypothesis. If our data is not extremely
unlikely assuming the null hypothesis is true, we fail to reject the null 
hypothesis.

## Hypothesis Testing Steps

We will use a real example to illustrate the steps of hypothesis testing. 
Student cell phone use during exams (texting, searching the internet, 
accessing class resources, etc) is a serious breach of academic integrity. 
Suppose you are conducting a study of cheating in college courses and are 
interested in determining if more than 5\% of students in a population of 
interest have used their cell phone during an exam.

#### 1. State the null and alternative hypothesis

The null hypothesis is denoted $H_o$ and represents the status quo claim that 
nothing unusual is going on. The alternative hypothesis is denoted $H_a$ and 
states the opposite - there is something going on that merits further 
investigation. Both hypotheses are defined for parameters (not statistics).

Consider the null and alternative hypotheses in words below.

- $H_o$: The true proportion of students who have used a cell phone during an 
exam is equal to 5\%.
- $H_a$: The true proportion of students who have used a cell phone during an 
exam is greater than 5\%.

```{r quiz1, echo=FALSE}
quiz(
  question("What is the appropriate way to express the null and alternative 
           hypothesis in symbols?",
    answer("Ho: p = 0.05 and Ha: p > 0.05",
      correct = TRUE,
      message = ""
    ),
    answer("Ho: p > 0.05 and Ha: p = 0.05",
      message = ""
    ),
    answer("Ho: p = 0.05 and Ha: p < 0.05",
      message = ""
    ),
    answer("Ho: p = 0.05 and Ha: p != 0.05",
      message = ""
    ),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

#### 2. Take a representative sample of the population, collect data, and summarize the data with a statistic.

Researchers took a representative sample of 267 students, surveyed them 
confidentially and found that 27 had used a cell phone during an exam. 

We first create a tibble with the sample data.

```{r}
cheating_data <-
  tibble(outcome = c(rep("text", 27), rep("no text", 267 - 27)))
glimpse(cheating_data)

cheating_data %>%
  count(outcome)
```


```{r quiz2, echo=FALSE}
quiz(
  question("The choice of summary statistic depends on the context. In this 
  example, what is a natural choice?",
    answer("27/267 = 0.1011",
      correct = TRUE,
      message = ""
    ),
    answer("1/20 = 0.05",
      message = ""
    ),
    answer("267/27",
      message = ""
    ),
    answer("1/2 = 0.50",
      message = ""
    ),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

#### 3. Determine how likely it is to see data as or more surprising as what we observed, assuming the null hypothesis is true.

We next find the probability of observing data as or more extreme as what we
observed, assuming $H_o$ is true. This is called the $p$-value.

First, we need to know what values of $\hat{p}$ we expect assuming the null 
hypothesis is true. In other words, we need to simulate a distribution for 
$\hat{p}$ assuming the probability of cheating is $0.05$.

This null distribution for $\hat{p}$ tells us what sample proportions to expect, 
assuming the null hypothesis is true.

We will generate the null distribution using the `infer` package. We must 
provide the following information.

1. `specify()` 
  - `response`: the outcome of interest in the appropriate data frame
  - `success`: the level of the outcome we are interested in
2. `hypothesize()`
  - `null`: null hypothesis (point if you are doing inference on a point estimate
  or independence if the null assumes independence between two variables)
  - `p`: appropriate probabilities of success and failure
3. `generate()` 
  - `reps`: how many resamples to generate
  - `type`: simulate for hypothesis testing (we did bootstrap earlier)
4. `calculate()` our sample statistic

Fill in the blanks in the code chunk below to generate the null distribution.

```{r ex1, exercise = TRUE}
set.seed(11)
null_dist <- ___ %>%
  specify(response = ___, success = ___) %>% 
  hypothesize(
    null = ___, 
    p = ___ 
  ) %>% 
  generate(reps = ___, type = ___) %>% 
  calculate(stat = ___)
```

```{r ex1-solution}
set.seed(11)
null_dist <- cheating_data %>%
  specify(response = outcome, success = "text") %>%
  hypothesize(
    null = "point",
    p = c("text" = 0.05, "no text" = 0.95)
  ) %>%
  generate(reps = 5000, type = "simulate") %>%
  calculate(stat = "prop")
```

We examine the first few rows of `null_dist`.

```{r echo = FALSE}
set.seed(11)
null_dist <- cheating_data %>%
  specify(response = outcome, success = "text") %>%
  hypothesize(
    null = "point",
    p = c("text" = 0.05, "no text" = 0.95)
  ) %>%
  generate(reps = 5000, type = "simulate") %>%
  calculate(stat = "prop")
```

```{r}
head(null_dist)
```

```{r quiz3, echo=FALSE}
quiz(
  question("Where should the null distribution be centered?",
    answer("0.05",
      correct = TRUE,
      message = ""
    ),
    answer("0.101",
      message = ""
    ),
    answer("0.50",
      message = ""
    ),
    answer("0",
      message = ""
    ),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

Use the `visualize` function in `infer` to visualize the null 
distribution.

```{r}
null_dist %>%
  visualize()
```

It remains to calculate the $p$-value. Recall we observed a sample proportion 
of $\hat{p} = 27 / 267 \approx 0.101$ and the $p$-value is the probability of
observing data as or more extreme as what we observed, assuming $H_o$ is true.

We will plot the null distribution again, but this time include a red vertical 
line at the observed proportion $\hat{p} = 27/267 = 0.101$.

```{r}
phat <- 27 / 267

null_dist %>%
  visualize() +
  shade_p_value(phat,
    direction = "greater"
  )
```

```{r quiz4, echo=FALSE}
quiz(
  question("Without calculating any numbers and only examining the plot above,
  is what  we observed in our sample surprising assuming the null hypothesis
  is true?",
    answer("Yes, we are extremely unlikely to observe sample proportions as 
           large or larger as what we observed assuming the null hypothesis is 
           true (p = 0.05).",
      correct = TRUE,
      message = ""
    ),
    answer("No, we are not surprised by sample proportions as large or larger 
           as what we observed assuming the null hypothesis is
           true (p = 0.05)."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

Now let's find the $p$-value

```{r}
null_dist %>%
  filter(stat >= (27 / 267)) %>%
  summarize(p_value = n() / nrow(null_dist))
```

#### 4. Make a Decision

We reject $H_o$ if the $p$-value is small. A small $p$-value means we are 
unlikely to observe something as or more extreme as what we observed
assuming $H_o$ is true.

How small is small enough? We use a threshold value called the significance or 
$\alpha$ level set prior to collecting data. A common choice is $\alpha = 0.05$. 
If $H_o$ is true, we expect to make the wrong decision 5\% of the time.

If the $p$-value is $\alpha$ or greater, the results are not statistically
significant.

```{r quiz5, echo=FALSE}
quiz(
  question("Based on the p-value, what is the appropriate statistical decision 
           and conclusion in context?",
    answer("We reject the null hypothesis. There is convincing evidence that 
           the proportion of college students who use their cell phones to 
           cheat on exams is greater than 0.05.",
      correct = TRUE,
      message = ""
    ),
    answer("We fail to reject the null hypothesis. There is convincing evidence 
    that the proportion of college students who use their cell phones to cheat 
    on exams is greater than 0.05."),
    answer("We reject the null hypothesis. We do not have convincing 
        evidence that the proportion of college students who use their cell 
        phones to cheat on exams is greater than 0.05."),
    answer("We fail to reject the null hypothesis. We do not have convincing 
        evidence that the proportion of college students who use their cell 
        phones to cheat on exams is greater than 0.05."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Your Turn!

We will use the `gss` data from the `infer` package. This data set contains a 
sample of 500 entries from the General Social Survey (GSS). The GSS is a 
national survey of American society and opinions that has been conducted since
1972.

First, familiarize yourself with the data set and variables using the code 
below. You can learn more about the General Social Survey by visiting their 
[website](https://gss.norc.org/).

We will examine the variable `sex`, the respondent's self-identified sex and are
interesting if the proportion of females in the population differs from 0.50.

First, find the observed statistic. You can do this using `dplyr` or using `specify` and `calculate` from `infer`.

```{r ex2, exercise = TRUE}

```

```{r ex2-solution}
phat <- gss %>%
  specify(response = sex, success = "female") %>%
  calculate(stat = "prop") %>%
  pull(stat)
```

Next, generate the null distribution.

```{r ex3, exercise = TRUE}
set.seed(42)

```

```{r ex3-solution}
set.seed(42)
null_dist <- gss %>%
  specify(response = sex, success = "female") %>%
  hypothesize(
    null = "point",
    p = c("female" = 0.50, "male" = 0.50)
  ) %>%
  generate(reps = 5000, type = "simulate") %>%
  calculate(stat = "prop")
```

Visualize the null distribution using the `visualize` function.

```{r ex4, exercise = TRUE}

```

```{r ex4-solution}
null_dist %>%
  visualize()
```

Starting with the plot above, add a vertical line representing the sample 
statistic and shade the region corresponding to the $p$-value.

```{r ex5, exercise = TRUE}

```

```{r ex5-solution}
null_dist %>%
  visualize() +
  shade_p_value(phat, direction = "two-sided")
```

```{r quiz6, echo=FALSE}
quiz(
  question("Based on the p-value, what is the appropriate statistical decision 
           and conclusion in context?",
    answer("We reject the null hypothesis. There is convincing evidence that 
           the proportion of Americans who report their sex as female is 
           different from 0.50.",
      message = ""
    ),
    answer("We fail to reject the null hypothesis. There is convincing evidence
    that  the proportion of Americans who report their sex as female is 
           different from 0.50."),
    answer("We reject the null hypothesis. We do not have convincing evidence that 
           the proportion of Americans who report their sex as female is 
           different from 0.50."),
    answer("We fail to reject the null hypothesis. We do not have convincing evidence that 
           the proportion of Americans who report their sex as female is 
           different from 0.50.",
      correct = TRUE
    ),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Resources and References

- [Infer Website](https://infer.netlify.app/)
- "The Use and Abuse of Cell Phones and Text Messaging During Class: A Survey of
College Students". College Teaching, 2012.
- [General Social Survey Website](https://gss.norc.org/)

## Return to Learn R Tutorials Page

[Click here.](https://duke-learning-r.netlify.app)